Detection and replication of epistasis influencing transcription in humans

### 2013-06-12 docs/manuscript/nature_letter.tex
added coverletter, sent copy of manuscript to big people

\title{Detection and replication of epistasis influencing transcription in humans}
\date{}
\author{Gibran Hemani, Konstantin Shakhbazov, Joseph E Powell}

@@

\begin{abstract}
A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Carlborg2004, Hill2008a, Crow2010}. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, multiple instances of epistasis are found. In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 238 genes. Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing. Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$). We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs. For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans that replicate in independent populations.
\end{abstract}

@@

% \subsection{Introduction}
The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits. Typically, SNPs are treated using an additive model where they are assumed to contribute linearly, independently, and cumulatively to the mean of a trait. This has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}. Outside the prism of human association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}. Here we demonstrate that amongst common polymorphisms multiple instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.

Detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009}. When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes. Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that many genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.

@@

We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).

@@

% \subsection{Discussion}
Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants. The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation. Further research into such genetic effects may provide a useful portal to understanding molecular mechanisms with greater clarity. With data and computational techniques now widely available the search for epistasis in larger datasets for traits of broader interest is warranted.

>>>>

\title{Detection and replication of epistasis influencing transcription in humans}
\date{}
\author{Gibran Hemani, Joseph E Powell, Konstantin Shakhbazov, \\
Greg Gibson, Grant W Montgomery, Peter M Visscher}

@@

\begin{abstract}
A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010}. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found. In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between single nucleotide polymorphisms (SNPs) acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing. Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects ($p < 1.0 \times 10^{-16}$). There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans.

\end{abstract}

@@

% \subsection{Introduction}
In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently, and cumulatively to the mean of a trait. This framework has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}. Outside the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes. Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that many genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.

@@

We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.

@@

% \subsection{Discussion}
Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants. The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation. Further research into such epistatic effects may provide a useful portal to understanding molecular mechanisms and complex trait variation with greater clarity. With computational techniques and data now widely available the search for epistasis in larger datasets for traits of broader interest is warranted.
